### 2020-07-10 submission/manuscript.tex
Reorganization edits

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of source.} To test how the
microbiotas of mice from different sources impact colonization dynamics
after clindamycin exposure, we utilized C57BL/6 mice from 6 different
sources: two colonies from the University of Michigan that were split
from each other 10 years ago (the Young and Schloss lab colonies) and
four commercial source: the Jackson Laboratory, Charles River
Laboratories, Taconic Biosciences, and Envigo (which was formerly
Harlan). These 4 vendors were chosen because they represent commonly
used vendors for CDI studies in mice (26, 34--40).

After a 13-day acclimation period for the mice ordered from vendors, all
mice were treated with 10 mg/kg clindamycin via intraperitoneal
injection and one day later challenged with 10\textsuperscript{3}
\emph{C. difficile} 630 spores (Fig. 1A). Clindamycin was chosen because
we have previously demonstrated mice are rendered susceptible, but
consistently cleared the CDI within 9 days (21, 41), clindamycin is
frequently implicated with human CDIs (42), and is also part of the
antibiotic treatment for the frequently cited 2008 CDI mouse model (14).
The day after infection, \emph{C. difficile} was detectable in all mice
at a similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 1B). Interestingly,
variation in \emph{C. difficile} CFU levels across sources of mice
emerged from days 3-7 post-infection (all \emph{P}\textsubscript{FDR}
\(\le\) 0.019; Fig. 1B and Table S1), suggesting mouse source is
associated with \emph{C. difficile} clearance. We conducted two
experiments approximately 3 months apart and while the colonization
dynamics were similar across most sources of mice, there was some
variation between the 2 experiments, particularly for the Schloss and
Envigo mice (Fig. S1A-B). Although \emph{C. difficile} 630 causes mild
symptoms in mice comparied to other \emph{C. difficile} strains (43), we
also saw that weight change significantly varied across sources of mice
with the most weight lost two days post-infection (Fig. 1C and Table
S2). Importantly, there was also one Jackson and one Envigo mouse that
died between 1- and 3-days post-infection during the second experiment.
Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
lose more weight (although there was variation between experiments with
Schloss and Envigo mice), have higher \emph{C. difficile} CFU levels and
take longer to clear the infection compared to the other sources of
mice, which was particularly evident 7 days post-infection (Fig. 1B-C,
Fig. S2C-D), when 57\% of the mice were still colonized with \emph{C.
difficile} (Fig. S1E). By 9 days post-infection the majority of the mice
from all sources had cleared \emph{C. difficile} (Fig. S1C) with the
exception of 1 Taconic mouse from the first experiment and 2 Envigo mice
from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Bacterial communities consistently vary across sources despite
antibiotic and infection perturbations.} Given the well known variation
in mouse microbiomes across breeding colonies (25), we hypothesized that
the variation in \emph{C. difficile} clearance could be explained by
microbiota variation across the 6 sources of mice. We used 16S rRNA gene
sequencing to characterize the fecal bacterial communities from the mice
over the course of the experiment. Since antibiotics and other risk
factors of CDIs are associated with decreased microbiota diversity (44),
we first examined alpha diversity measures across the 6 sources of mice.
Examining the bacterial communities at baseline, prior to clindamycin
treatment, there was a significant difference in the number of observed
OTUs (\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity
index (\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig.
2A-B and Table S3). As expected, the clindamycin treatment decreased
richness and Shannon diversity across all sources of mice, and richness
and Shannon diversity started to increase 1 day post-infection (Fig.
2C-D). Interestingly, significant differences in diversity metrics
across sources (\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged
after both clindamycin and \emph{C. difficile} infection, with Charles
River mice having higher richness and Shannon Diversity than most of the
other sources (Fig 2C-F and Table S4). Although the Charles River mice
had more diverse microbiotas, the Young and Schloss lab mice were also
able to clear \emph{C. difficile} faster than the other sources,
suggesting microbiota diversity alone does not explain the observed
variation in \emph{C. difficile} colonization across vendors.

Next, we compared the community structure of mice from the 6 sources
over the course of the experiment using principal coordinate analysis
(PCoA) of the \(\theta_{YC}\) distances. Permutational multivariate
analysis of variance (PERMANOVA) analysis revealed source was the
dominant factor that explained the observed variation across fecal
communities (R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage (mice from the same sources were housed
together, primarily at a density of 2 mice per cage) and day of the
experiment (Movie S1 and Table S5). Mice that are co-housed have been
shown to have similar gut microbiotas due to coprophagy (45) and since
mice within the same source were housed together, it is not surprising
that cage also contributes to the observed microbiota variation. Since
the majority of the perturbations happened over the initial days of the
experiment, we decided to focus on the communities at baseline (day -1),
after clindamycin treatment (day 0), and post-infection (day 1). For all
3 timepoints, the source and cage interaction significantly explained
most of the observed community variation (combined R\textsuperscript{2}
= 0.90, 0.99, 0.88, respectively; \emph{P} = 0.0001; Fig. 3 and Table
S6). We also compared baseline communities across the 2 experiments, and
found experiment and cage significantly explained the observed variation
only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table
S7). However, most of the vendors also clustered by experiment (Fig.
S2C-D, F), suggesting there was some community variation between the 2
experiments within each source. Thus, source was the factor that
explained the most variation observed in the bacterial communities.
Importantly with the exception of the Schloss and Young colonies, the
community of each source clustered apart from one another suggesting
each community had a unique response to clindamycin treatment and
\emph{C. difficile} challenge.

Since there was some variation in the microbiota structure of mice from
the same source between experiments at baseline, we next looked at how
similar the communities were within the same source and between sources
in response to clindamycin treatment and \emph{C. difficile} challenge
(Fig. 4) . The baseline communities varied most between experiments for
Schloss, Young, and Envigo mice and variation between sources of mice
was high (Fig. 4A). Clindamycin treatment reduced the variation between
experiments within Schloss, Young and Jackson mice and some of the
variation between sources diminished, particularly for the Schloss,
Young, and Charles River mice (Fig. 4B). Post-infection, the community
variation started to increase within sources of mice and variation
between sources of mice started to increase towards previous levels
(Fig. 4C). By using mice from multiple sources we were able to increase
the number of microbiota communities we tested with the clindamycin
\emph{C. difficile} colonization mouse model.

After finding differences at the community level, we next identified the
bacteria that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) level. Focusing on the baseline communities first,
there were 268 OTUs (Table S8) with relative abundances that varied
across sources. Clindamycin treatment reduced the number of taxa with
relative abundances that varied across sources to 18 OTUs (Table S8).
After \emph{C. difficile} challenge, there were 44 OTUs (Table S8) with
significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 5A-C) with relative abundances that consistently
varied across sources. Importantly, some of the OTus that consistently
varied across sources also shifted with clindamycin treatment. For
example, \emph{Proteus} increased after clindamycin treatment, but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased after clindamycin
treatment. In summary, mouse bacterial communities varied significantly
according to source throughout the course of the experiment and a
consistent subset of bacterial taxa remained different across sources
regardless of clindamycin and \emph{C. difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all sources.} Although there were bacteria that consistently varied
across sources, we also wanted to identify the bacteria that shifted
after clindamycin treatment, regardless of source. By analyzing all mice
that had sequence data from fecal samples collected at baseline and
after clindamycin treatment, we identified 153 OTUs that were altered
after clindamycin treatment (Fig. 6 and Tables S9). Interestingly, when
we compared the list of significant clindamycin impacted bacteria with
the bacteria that consistently varied across groups over the initial 3
timepoints of our experiment, we found 3 OTUs (\emph{Lachnospiraceae}
(OTU 130), \emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23))
overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that
clindamycin has a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs also varied
across sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by source, changed after clindamycin treatment, or both, we
next wanted to determine which taxa were influencing the variation in
\emph{C. difficile} colonization at day 7 (Fig. 1D, Fig. S1C). We
trained three L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3A-B). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5.2e-31; Table S12).
Interestingly, the model based on the post-clindamycin (day 0) community
OTU data performed significantly better than all other models with an
area under the receiving operator characteristic curve (AUROC) of 0.75
(\emph{P}\textsubscript{FDR} \(\le\) 3.1e-11 for pairwise comparisons;
Table S13). Thus, we were able to use bacterial relative abundance data
alone to differentiate mice that had cleared \emph{C. difficile} before
day 7 from the mice still colonized with \emph{C. difficile} at that
timepoint. Interestingly, the model built with OTU relative abundance
data post-clindamycin treatment had the best performance, suggesting how
the bacterial community responds to clindamycin treatment has the
greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Tables S14-15). First, we
looked at OTUs from the model with the best performance that was based
on the post-clindamycin treatment bacterial community data. While most
of the 20 OTUs had low relative abundances on day 0,
\emph{Enterobacteriace}, \emph{Bacteroides} and \emph{Proteus} had high
relative abundances in at least one source of mice and significantly
varied across sources (Fig. 7A). Next, the top 20 taxa from each model
were compared to the list of taxa that varied across source at the same
timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by
clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of
OTUs that were important to the model and overlapped with bacteria that
varied by either source, clindamycin treatment, or both (Fig. S4,
S5A-C). Combining the overall results for the 3 OTU models identified 14
OTUs associated with source, 21 OTUs associated with clindamycin
treatment, and 6 OTUs associated with both (Fig. 7B). Several OTUs
(\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
(OTU 1), Porphyromonadaceae (OTU 7)}) appeared across at least 2 models,
so we examined how the relative abundances of these key taxa varied over
the course of the experiment (Fig. 8). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these OTUs
significantly varied across sources (Table S13). Interestingly, there
were no OTUs that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection with \emph{C. difficile}
630, suggesting multiple bacteria influence the time needed to clear the
infection. Together, these results suggest the initial bacterial
communities and their responses to clindamycin influence the clearance
of \emph{C. difficile}.

@@

By running our CDI model with mice from 6 different sources, we were
able to identify bacterial taxa that were unique to sources throughout
the experiment as well as taxa that were universally impacted by
clindamycin. We trained L2 logistic regression models with baseline,
post-clindamycin treatment, and post-infection fecal community data that
could predict whether mice cleared \emph{C. difficile} by 7 days
post-infection better than random chance. We identified
\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 8) as candidate bacteria
within these communities that were influencing variation in \emph{C.
difficile} colonization dynamics since these bacteria were all important
in the logistic regression models and varied by source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrate
clindamycin is sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

@@

Differences in CDI mouse model studies have been attributed to different
intestinal microbiotas of mice from different sources. For example,
groups using the same clindamycin treatment and C57BL/6 mice had
different \emph{C. difficile} outcomes, one having sustained
colonization (32), while the other had transient (18), despite both
using \emph{C. difficile} VPI 10643. Baseline differences in the
microbiota composition have been hypothesized to partially explain the
differences in colonization outcomes and overall susceptibility to
\emph{C. difficile} after treatment with the same antibiotic (13, 31).
The bacterial perturbations induced by clindamycin treatment have been
well characterized and our findings agree with previous CDI mouse model
work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
were associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 41, 46--48). While we have demonstrated that
susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments. The \emph{C.
difficile} strain used could also be contributing to the variation in
\emph{C. difficile} outcomes seen across different research groups (47).
We found the time needed to naturally clear \emph{C. difficile} varied
across sources of mice implying that at least in the context of the same
perturbation, microbiota differences seemed to influence infection
outcome more than susceptibility. More importantly, we were able to
reduce the variation observed across sources to identify a subset of
OTUs that were also important for predicting \emph{C. difficile}
colonization status 7 days post-infection. Since all but 3 mice
eventually cleared \emph{C. difficile} 630 by 9 days post-infection and
the model built with the post-clindamycin OTU relative abundance data
had the best performance, our results suggest clindamycin treatment had
a large role in determining \emph{C. difficile} susceptibility and
clearance in the mice.

@@

There are several limitations to our work. The microbiome is composed of
viruses, fungi, and parasites in addition to bacteria, and these
non-bacterial members can also vary across mouse vendors (57, 58). While
our study focused solely on the bacterial portion, viruses and fungi
have also begun to be implicated in the context of CDIs or FMT
treatments for recurrent CDIs (35, 59--62). Beyond community
composition, the metabolic function of the microbiota also has a CDI
signature (20, 48, 63, 64) and can vary across mice from different
sources (65). For example, microbial metabolites, particularly secondary
bile acids and butyrate production, have been implicated as important
contributors to \emph{C. difficile} resistance (33, 47). Although, we
only looked at composition, \emph{Ruminococcaceae} and
\emph{Lachnospiraceae} both emerged as important taxa for classifying
day 7 \emph{C. difficile} colonization status and metagenomes from these
bacteria have been shown to contain the bile acid-inducible gene cluster
necessary for secondary bile acid formation and ability to produce
butyrate (52, 66). Interestingly, butyrate has previously been shown to
vary across vendors and mediated resistance to \emph{Citrobacter
rodentium} infection, a model of enterohemorrhagic and enteropathogenic
\emph{Escherichia coli} infections (23). Evidence for immunological
toning differences in IgA and Th17 cells across mice from different
vendors have also been documented and (67, 68) may also influence
response to CDI, particularly in the context of severe CDIs (69, 70).
The outcome after \emph{C. difficile} exposure depends on a multitude of
factors, including age, diet, and immunity; all of which are also
influenced by the microbiota.

@@

\textbf{(vii) Statistical analysis.} All statistical tests were
performed in R (v 3.5.2) (78). The Kruskal-Wallis test was used to
analyze differences in \emph{C. difficile} CFU, mouse weight change, and
alpha diversity across vendors with a Benjamini-Hochberg correction for
testing multiple timepoints, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. For taxonomic analysis and
generation of logistic regression model input data, \emph{C. difficile}
(OTU 20) was removed. Bacterial relative abundances that varied across
sources at the OTU level were identified with the Kruskal-Wallis test
with Benjamini-Hochberg correction for testing all identified OTUs,
followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
correction. OTUs impacted by clindamycin treatment were identified using
the Wilcoxon signed rank test with matched pairs of mice samples for day
-1 and day 0. To determine whether classification models had better
performance (test AUROCs) than random chance (0.5), we used the
one-sample Wilcoxon signed rank test. To examine whether there was an
overall difference in predictive performance across the 6 classification
models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
comparisons with Benjamini-Hochberg correction for multiple hypothesis
testing. The tidyverse package was used to wrangle and graph data (v
1.3.0) (81).

@@

\hypertarget{ref-Tomkovich2019}{}
41. Tomkovich S, Lesniak NA, Li Y, Bishop L, Fitzgerald MJ, Schloss PD.
2019. The proton pump inhibitor omeprazole does not promote
\emph{Clostridioides difficile} colonization in a murine model. mSphere
4.

\hypertarget{ref-Guh2018}{}
42. Guh AY, Kutty PK. 2018. \emph{Clostridioides difficile} infection
169:ITC49.

>>>>

\textbf{Microbiota variation is high between mice from different
sources.} To test how the microbiotas of mice from different sources
impact colonization dynamics after clindamycin exposure, we utilized
C57BL/6 mice from 6 different sources: two colonies from the University
of Michigan that were split from each other 10 years ago (the Young and
Schloss lab colonies) and four commercial source: the Jackson
Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo
(which was formerly Harlan). These 4 vendors were chosen because they
represent commonly used vendors for CDI studies in mice (26, 34--40). We
used 16S rRNA gene sequencing to characterize the fecal bacterial
communities across sources at baseline.

Since antibiotics and other risk factors of CDIs are associated with
decreased microbiota diversity (41), we first examined alpha diversity
across the 6 sources of mice. Intially, there was a significant
difference in the number of observed operational taxonomic units (OTUs),
but not Shannon diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively) across sources of
mice (Fig. 1A-B and Table S1). Next, we compared the community structure
of mice from the 6 sources over the course of the experiment using
principal coordinate analysis (PCoA) of the \(\theta_{YC}\) distances.
Permutational multivariate analysis of variance (PERMANOVA) analysis
revealed source and cage interactions explained most of the observed
variation across fecal communities explained most of the observed
community variation (combined R\textsuperscript{2} = 0.90, \emph{P} =
0.0001, Fig. 3 and Table S2). Mice that are co-housed have been shown to
have similar gut microbiotas due to coprophagy (42) and since mice
within the same source were housed together, it is not surprising that
cage also contributed to the observed microbiota variation. Since we
conducted two experiments approximately 3 months apart, we also compared
baseline communities across the 2 experiments. Experiment and cage
significantly explained the observed variation only for the Schloss and
Young lab mouse colonies (Fig. S2A-B and Table S2). However, most of the
vendors also clustered by experiment (Fig. S2C-D, F), suggesting there
was some community variation between the 2 experiments within each
source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
After finding differences at the community level, we next identified the
bacteria that initially varied across sources of mice. There were 268
OTUs with relative abundances that varied across sources (Fig. 1D and
Table S3). By using mice from six sources we were able to increase the
number of microbiota communities to test with the clindamycin \emph{C.
difficile} colonization mouse model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
After the acclimation period, all mice were treated with 10 mg/kg
clindamycin via intraperitoneal injection and one day later challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 2A).
Clindamycin was chosen because we have previously demonstrated mice are
rendered susceptible, but consistently cleared \emph{C. difficile}
within 9 days (21, 43), clindamycin is frequently implicated with human
CDIs (44), and is also part of the antibiotic treatment for the
frequently cited 2008 CDI mouse model (14). The day after infection,
\emph{C. difficile} was detectable in all mice at a similar level
(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
indicating clindamycin rendered all mice susceptible regardless of
source (Fig. 2B). Interestingly, variation in \emph{C. difficile} CFU
levels across sources of mice emerged from days 3-7 post-infection (all
\emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 2B and Table S4),
suggesting mouse source is associated with \emph{C. difficile}
clearance. We conducted two experiments and while the colonization
dynamics were similar across most sources of mice, there was some
variation between the 2 experiments, particularly for the Schloss and
Envigo mice (Fig. S2A-B). Although \emph{C. difficile} 630 causes mild
symptoms in mice comparied to other \emph{C. difficile} strains (45), we
also saw that weight change significantly varied across sources of mice
with the most weight lost two days post-infection (Fig. 2C and Table
S5). Importantly, there was also one Jackson and one Envigo mouse that
died between 1- and 3-days post-infection during the second experiment.
Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to
lose more weight, have higher \emph{C. difficile} CFU levels and take
longer to clear the infection compared to the other sources of mice
(although there was variation between experiments with Schloss and
Envigo mice), which was particularly evident 7 days post-infection (Fig.
2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
the mice from all sources had cleared \emph{C. difficile} (Fig. 2C) with
the exception of 1 Taconic mouse from the first experiment and 2 Envigo
mice from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the well
known variation in mouse microbiomes across breeding colonies (25), we
hypothesized that variation in \emph{C. difficile} clearance would be
explained by microbiota variation across the 6 sources of mice. We used
16S rRNA sequencing to characterize microbiota variation across sources
after clindamycin treatment and throughout the rest of the experiment.
PERMANOVA analysis revealed source was the dominant factor that
explained the observed variation across fecal communities
(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage (mice from the same sources were housed
together, primarily at a density of 2 mice per cage) and day of the
experiment (Movie S1 and Table S2). Since the majority of the
perturbations happened over the initial days of the experiment, we
decided to focus on the communities after clindamycin treatment (day 0)
and post-infection (day 1).

As expected, clindamycin treatment decreased richness and Shannon
diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
treatment, with Charles River mice having higher richness and Shannon
diversity than most of the other sources (Fig 3A-B and Table S1). The
variation in community structures was also diminished by clindamycin
treatment (Fig. 3C and Table S2). Next we identified the bacteria that
shifted after clindamycin treatment, regardless of source by analyzing
all mice that had sequence data from fecal samples collected at baseline
and after clindamycin treatment. We identified 153 OTUs that were
altered after clindamycin treatment (Fig. 3D and Table S6) in most mice.
Despite the significant impact of clindamycin, there were still 18 OTUs
(Fig. 3E and Table S3) with relative abundances that significantly
varied across sources. Interestingly, when we compared the list of
significant clindamycin impacted bacteria with the bacteria that varied
across sources post-clindamycin, we found 4 OTUs
(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
(Fig. 3D-E and Tables S3, S6). These findings demonstrate that
clindamycin has a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs continued to vary
across sources.

\textbf{Microbiota variation across sources is maintained after \emph{C.
difficile} challenge.} Post-infection, significant differences in
diversity metrics were maintained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Table S1).
Although the Charles River mice had more diverse microbiotas, the Young
and Schloss lab mice were also able to clear \emph{C. difficile} faster
than some of the other sources, suggesting microbiota diversity alone
does not explain the observed variation in \emph{C. difficile}
colonization across sources. Source and cage interaction continued to
explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88, respectively; \emph{P} = 0.0001; Fig. 4D
and Table S2). After \emph{C. difficile} challenge, there were 44 OTUs
(Fig. 4D and Table S3) with significantly different relative abundances
across sources. In spite of the experimental perturbations that
occurred, there were 12 OTUs with relative abundances that consistently
varied across sources at baseline, post-clindamycin, and post-infection.
Importantly, some of the OTUs that consistently varied across sources
also shifted with clindamycin treatment. For example, \emph{Proteus}
increased after clindamycin treatment, but only in Taconic mice.
\emph{Enterococcus} was primarily found only in mice purchased from
commercial vendors and also increased after clindamycin treatment. In
summary, mouse bacterial communities varied significantly according to
source throughout the course of the experiment and a consistent subset
of bacterial taxa remained different across sources regardless of
clindamycin and \emph{C. difficile} challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied by source, changed
after clindamycin treatment, or both, we next wanted to determine which
taxa were influencing the variation in \emph{C. difficile} colonization
at day 7 (Fig. 2D, Fig. S2C). We trained three L2-regularized logistic
regression models with input bacterial community data from the baseline,
post-clindamycin, and post-infection timepoints of the experiment to
predict \emph{C. difficile} colonization status on day 7 (Fig. S3A-B).
All models were better at predicting \emph{C. difficile} colonization
status on day 7 than random chance (all \emph{P} \(\le\) 1.6e-17; Table
S7). Interestingly, the model based on the post-clindamycin (day 0)
community OTU data performed significantly better than the other models
with an area under the receiving operator characteristic curve (AUROC)
of 0.78 (\emph{P}\textsubscript{FDR} \(\le\) 3.3e-06 for pairwise
comparisons; Table S7). Thus, we were able to use bacterial relative
abundance data alone to differentiate mice that had cleared \emph{C.
difficile} before day 7 from the mice still colonized with \emph{C.
difficile} at that timepoint. Interestingly, the model built with OTU
relative abundance data post-clindamycin treatment had the best
performance, suggesting how the bacterial community responds to
clindamycin treatment has the greatest influence on subsequent \emph{C.
difficile} colonization dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 OTUs that had the highest absolute
feature weights in each of the 6 models (Table S8). First, we looked at
OTUs from the model with the best performance that was based on the
post-clindamycin treatment bacterial community data. While most of the
20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
least one source of mice and significantly varied across sources (Fig.
5A). Next, the top 20 taxa from each model were compared to the list of
taxa that varied across source at the same timepoint (Fig. 1D, 3E, 4D
and Table S3) and the taxa that were altered by clindamycin treatment
(Fig. 3D and Table S6). We found a subset of OTUs that were important to
the model and overlapped with bacteria that varied by either source,
clindamycin treatment, or both (Fig. S4). Combining the overall results
for the 3 models identified 14 OTUs associated with source, 21 OTUs
associated with clindamycin treatment, and 6 OTUs associated with both
(Fig. 5B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU
23), Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) appeared
across at least 2 models, so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Throughout
the experiment, there was at least 1 timepoint where relative abundances
of these OTUs significantly varied across sources (Table S9).
Interestingly, there were no OTUs that emerged as consistently enriched
or depleted in mice that were colonized past 7 days post-infection with
\emph{C. difficile} 630, suggesting multiple bacteria influence the time
needed to clear the infection. Together, these results suggest the
initial bacterial communities and their responses to clindamycin
influence the clearance of \emph{C. difficile}.

@@

By running our CDI model with mice from 6 different sources, we were
able to identify bacterial taxa that were unique to sources throughout
the experiment as well as taxa that were universally impacted by
clindamycin. We trained L2 logistic regression models with baseline,
post-clindamycin treatment, and post-infection fecal community data that
could predict whether mice cleared \emph{C. difficile} by 7 days
post-infection better than random chance. We identified
\emph{Bacteroides (OTU 2), Enterococcus (OTU 23), Enterobacteriaceae
(OTU 1), Porphyromonadaceae (OTU 7)} (Fig. 6) as candidate bacteria
within these communities that were influencing variation in \emph{C.
difficile} colonization dynamics since these bacteria were all important
in the logistic regression models and varied by source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrate
clindamycin is sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

@@

Differences in CDI mouse model studies have been attributed to different
intestinal microbiotas of mice from different sources. For example,
groups using the same clindamycin treatment and C57BL/6 mice had
different \emph{C. difficile} outcomes, one having sustained
colonization (32), while the other had transient (18), despite both
using \emph{C. difficile} VPI 10643. Baseline differences in the
microbiota composition have been hypothesized to partially explain the
differences in colonization outcomes and overall susceptibility to
\emph{C. difficile} after treatment with the same antibiotic (13, 31).
The bacterial perturbations induced by clindamycin treatment have been
well characterized and our findings agree with previous CDI mouse model
work demonstrating \emph{Enterococcus} and \emph{Enterobacteriaceae}
were associated with \emph{C. difficile} susceptibility and
\emph{Porpyhromonadaceae}, \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, and \emph{Turicibacter} were associated with
resistance (19, 21, 32, 33, 43, 46--48). While we have demonstrated that
susceptibility is uniform across sources of mice after clindamycin
treatment, there could be different outcomes for either susceptibility
or clearance in the case of other antibiotic treatments. The \emph{C.
difficile} strain used could also be contributing to the variation in
\emph{C. difficile} outcomes seen across different research groups. For
example, a group found differential colonization outcomes after
clindamycin treatment, with \emph{C. difficile} 630 and M68 eventually
becoming undectable while BI-7 remained detectable up to 70 days
post-treatment (47). We found the time needed to naturally clear
\emph{C. difficile} varied across sources of mice implying that at least
in the context of the same perturbation, microbiota differences seemed
to influence infection outcome more than susceptibility. More
importantly, we were able to reduce the variation observed across
sources to identify a subset of OTUs that were also important for
predicting \emph{C. difficile} colonization status 7 days
post-infection. Since all but 3 mice eventually cleared \emph{C.
difficile} 630 by 9 days post-infection and the model built with the
post-clindamycin OTU relative abundance data had the best performance,
our results suggest clindamycin treatment had a large role in
determining \emph{C. difficile} susceptibility and clearance than the
source of the mice.

@@

There are several limitations to our work. The microbiome is composed of
viruses, fungi, and parasites in addition to bacteria, and these
non-bacterial members can also vary across mouse sources (57, 58). While
our study focused solely on the bacterial portion, viruses and fungi
have also begun to be implicated in the context of CDIs or FMT
treatments for recurrent CDIs (35, 59--62). Beyond community
composition, the metabolic function of the microbiota also has a CDI
signature (20, 48, 63, 64) and can vary across mice from different
sources (65). For example, microbial metabolites, particularly secondary
bile acids and butyrate production, have been implicated as important
contributors to \emph{C. difficile} resistance (33, 47). Although, we
only looked at composition, \emph{Ruminococcaceae} and
\emph{Lachnospiraceae} both emerged as important taxa for classifying
day 7 \emph{C. difficile} colonization status and metagenomes from these
bacteria have been shown to contain the bile acid-inducible gene cluster
necessary for secondary bile acid formation and ability to produce
butyrate (52, 66). Interestingly, butyrate has previously been shown to
vary across vendors and mediated resistance to \emph{Citrobacter
rodentium} infection, a model of enterohemorrhagic and enteropathogenic
\emph{Escherichia coli} infections (23). Evidence for immunological
toning differences in IgA and Th17 cells across mice from different
vendors have also been documented and (67, 68) may also influence
response to CDI, particularly in the context of severe CDIs (69, 70).
The outcome after \emph{C. difficile} exposure depends on a multitude of
factors, including age, diet, and immunity; all of which are also
influenced by the microbiota.

@@

\textbf{(vii) Statistical analysis.} All statistical tests were
performed in R (v 3.5.2) (78). The Kruskal-Wallis test was used to
analyze differences in \emph{C. difficile} CFU, mouse weight change, and
alpha diversity across sources with a Benjamini-Hochberg correction for
testing multiple timepoints, followed by pairwise Wilcoxon comparisons
with Benjamini-Hochberg correction. For taxonomic analysis and
generation of logistic regression model input data, \emph{C. difficile}
(OTU 20) was removed. Bacterial relative abundances that varied across
sources at the OTU level were identified with the Kruskal-Wallis test
with Benjamini-Hochberg correction for testing all identified OTUs,
followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg
correction. OTUs impacted by clindamycin treatment were identified using
the Wilcoxon signed rank test with matched pairs of mice samples for day
-1 and day 0. To determine whether classification models had better
performance (test AUROCs) than random chance (0.5), we used the
one-sample Wilcoxon signed rank test. To examine whether there was an
overall difference in predictive performance across the 6 classification
models we used the Kruskal-Wallis test followed by pairwise Wilcoxan
comparisons with Benjamini-Hochberg correction for multiple hypothesis
testing. The tidyverse package was used to wrangle and graph data (v
1.3.0) (81).

@@

\hypertarget{ref-Ross2016}{}
41. Ross CL, Spinler JK, Savidge TC. 2016. Structural and functional
changes within the gut microbiota and susceptibility to
\emph{Clostridium difficile} infection. Anaerobe 41:37--43.

\hypertarget{ref-Nguyen2015}{}
42. Nguyen TLA, Vieira-Silva S, Liston A, Raes J. 2015. How informative
is the mouse for human gut microbiota research? Disease Models \&
Mechanisms 8:1--16.
